---
title: "LAGOS Spatial Analysis"
author: "Matthew Ross | Assignment completed by: George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    # code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(viridis)
library(RColorBrewer)
library(scales)
# spatial analysis
library(RApiSerialize)
library(LAGOSNE)
library(USAboundaries)
library(sf)
library(mapview) #Interactive maps
library(leafpop) #map html popup

```

# LAGOS Analysis

## Loading in data

### First download and then specifically grab the locus (or site lat longs)

```{r, warning=F, message=F, results='hide'}
# #Lagos download script

# LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())

#Load in lagos
lagos <- lagosne_load()
names(lagos)

#Grab the lake centroid info
lake_centers <- lagos$locus
str(lake_centers)

```

### Convert to spatial data

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
#Look at the column names
#names(lake_centers)

#Look at the structure
#str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))

spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>% # EPSG:4326 = WGS 84
  st_transform(2163) # EPSG:2163 = US National Atlas Equal Area

#Subset for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

# subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
mapview(subset_spatial)

```

### Subset to only Minnesota

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
states <- us_states()

#Plot all the states to check if they loaded
#mapview(states)

minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163) # EPSG:2163 = US National Atlas Equal Area

#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,]

#Plotting the first 1000 lakes
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha')
```


# In-Class work

## Question 1)
Show a map outline of Iowa and Illinois (similar to Minnesota map upstream)

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}

#filter states dataset from USA USAboundaries us_states()
il_ia <- states %>%
  filter(name %in% c("Illinois", "Iowa")) %>%
  st_transform(2163) # EPSG:2163 = US National Atlas Equal Area

#see names of basemaps available to pass to "map.types" option below
# names(leaflet.providers::providers_loaded()$providers)

#make map
mapview(il_ia
        , zcol = "name"
        , legend = FALSE
        , col.regions = turbo(n= min(5, n_distinct(il_ia$name)))
        , alpha.regions = 0.5
        , map.types =  "OpenStreetMap"
        , label = c("stusps")
    )

```


## Question 2)
Subset LAGOS data to these sites, how many sites are in Illinois and Iowa
combined? How does this compare to Minnesota?

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
#Subset lakes based on spatial position
il_ia_lakes <- spatial_lakes[il_ia,]

#create counts dataset
cnt_lakes <- c(
  n_distinct(il_ia_lakes$lagoslakeid)
  , n_distinct(minnesota_lakes$lagoslakeid)
)

st <- c("Iowa + Illinois", "Minnesota")

df <- data.frame(cnt_lakes, st)

#plot
ggplot(df, aes(x = st, y = cnt_lakes, fill = st)) +
  geom_col(alpha=0.85, width = 0.5) + 
  geom_text(aes(label = scales::comma(cnt_lakes)), colour = "black", size = 4, vjust = "top") +
  scale_y_continuous(labels = label_comma()) +
  labs(
      title = "Iowa + Illinois Lakes vs. Minnesota Lakes"
    ) +
  xlab("") +
  ylab("# Lakes") +
  scale_fill_viridis(discrete = TRUE, option = "cividis") +
  theme_bw() +
  theme(
    legend.position="none"
    , axis.text.y = element_text(size = 11)
    , axis.text.x = element_text(size = 11)
  ) 

```

<span style="color: blue;">There are **`r scales::comma(n_distinct(il_ia_lakes$lagoslakeid))`** lakes in Iowa and Illinois combined while Minnesota has **`r scales::comma(n_distinct(minnesota_lakes$lagoslakeid))`** lakes. There are **`r scales::comma(n_distinct(minnesota_lakes$lagoslakeid) - n_distinct(il_ia_lakes$lagoslakeid))`** more lakes in Minnesota than in Iowa and Illinois combined.</span>


## Question 3)
What is the distribution of lake size in Iowa vs. Minnesota? Here I want to see a histogram plot with lake size on x-axis and frequency on y axis (check out geom_histogram)

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
# get Minnesota and Iowa polygons
mn_ia <- states %>%
  filter(name %in% c("Minnesota", "Iowa")) %>%
  st_transform(2163) # EPSG:2163 = US National Atlas Equal Area

# spatial join points to polygons
mn_ia_lakes <- st_join(spatial_lakes, mn_ia, left = FALSE)

# count lakes
# n_distinct(mn_ia_lakes$lagoslakeid)

#plot
ggplot(mn_ia_lakes, aes(lake_area_ha, fill = name)) +
  geom_histogram(bins = 15, alpha = 0.8) +
  facet_grid(. ~ name) +
  scale_y_continuous(labels = label_comma()) +
  labs(
      title = "Distribution of Lake Area in Minnesota and Iowa"
    ) +
  xlab("Lake Area (ha)") +
  ylab("# Lakes") +
  scale_fill_viridis(discrete = TRUE, option = "cividis") +
  theme_bw() +
  theme(
    legend.position="none"
    , axis.text.y = element_text(size = 11)
    , axis.text.x = element_text(size = 11)
  ) 

```

<span style="color: blue;">This histogram is not very informative due to outliers at the high end of lake area. Below is a quick solution to bin the top 5% of lakes based on area into their own bucket. In addition, to get a better comparison between states, we can use the `after_stat(density)` option to plot density on the y-axis instead of count. </span>


```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
# truncate top 5% of lakes in area into own bin
mn_ia_lakes <- mn_ia_lakes %>% 
  mutate(lake_area_ha_trunc = ifelse(
      lake_area_ha >= quantile(mn_ia_lakes$lake_area_ha, 1-0.05)
      , quantile(mn_ia_lakes$lake_area_ha, 1-0.05)
      , lake_area_ha)
  )

# plot
ggplot(mn_ia_lakes, aes(lake_area_ha_trunc, after_stat(density), fill = name)) +
  geom_histogram(bins = 30, alpha = 0.8) +
  facet_grid(. ~ name) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
      title = "Distribution of Lake Area in Minnesota and Iowa"
    ) +
  xlab("Lake Area (ha)") +
  ylab("% Lakes") +
  scale_fill_viridis(discrete = TRUE, option = "cividis") +
  theme_bw() +
  theme(
    legend.position="none"
    , axis.text.y = element_text(size = 11)
    , axis.text.x = element_text(size = 11)
    , strip.text.x = element_text(size = 13)
  ) 

```

<span style="color: blue;">This histogram is more informative and allows us to see that Iowa has a higher propoertion of small lakes compared to Minnesota. There are many more large lakes (over **~`r round(quantile(mn_ia_lakes$lake_area_ha, 1-0.05))`** ha) in Minnesota than in Iowa.</span>


## Question 4)
Make an interactive plot of lakes in Iowa and Illinois and color them
by lake area in hectares

```{r, warning=F, message=F, fig.width = 8, fig.height = 5}
#sort lakes so that largest lakes are on bottom
il_ia_lakes_filter <- il_ia_lakes %>% 
  arrange(-lake_area_ha) %>% 
  slice(1:1000) %>% 
  mutate(lake_area_ha_log = log10(lake_area_ha))

#point map
mapview(il_ia_lakes_filter
          , zcol = "lake_area_ha"
          , cex = "lake_area_ha_log"
          , alpha.regions = 0.5
          , map.types =  "OpenStreetMap"
          , label = c("gnis_name")
          , layer.name = "lake area (ha)"
          , popup = popupTable(
              il_ia_lakes_filter
              , zcol = c(
                "gnis_name"
                , "lake_area_ha"
                , "lagoslakeid"
              )
              , row.numbers = FALSE
              , feature.id = FALSE
            )
      )

```

<span style="color: blue;">The interactive map of lakes is displaying only the largest 1000 lakes in Illinois and Iowa. Because there were **`r scales::comma(n_distinct(il_ia_lakes$lagoslakeid))`** lakes in the two states, a filter is helpful in limiting the noise in the plot from too many points.</span>

## Question 5)
What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 

<span style="color: blue;">The r package ['LAGOSNE'](https://cran.r-project.org/web/packages/LAGOSNE/LAGOSNE.pdf) aids users in accessing data from the [Lake Multi-Scaled Geospatial and Temporal Database](https://lagoslakes.org/). In addition to the `locus` dataset utilized above, the package also contains the `lakes_limno` dataset which includes measures of mean and maximum lake depth (meters). The package also makes the following datasets available:</span>

<span style="color: blue;">* `chag` datasets contain information on climate, hydrology, atmospheric deposition, and surficial geology characteristics.</span>

<span style="color: blue;">* `conn` datasets contain lake, stream, and wetland abundance and connectivity metrics.</span>

<span style="color: blue;">* `lulc` datasets contain information on the land use and land cover (LULC) characteristics.</span>

## Link to part 2
[Go to part 2](https://georgewoolsey.github.io/LAGOS_Spatial_Analyses/part2#Class_work)